GVN-Hoist: Hoisting Computations from Branches

### 2016-08-08 hoist.tex
more on intro

This paper describes code hoisting, a technique that uses the information
computed for PRE to detect identical computations and has a transformation phase
whose goal differs from PRE: it removes identical computations from different
branches of execution.  These identical computations would not be redundant at
run-time and thus the number of run-time computations is not reduced.  Instead,
the goals of code hoisting are:
\begin{itemize}
\item to reduce the code size of the program,
\item to expose more instruction level parallelism by hoisting identical
  computations to be executed earlier, and thus helping to schedule heavy
  computations earlier in order to avoid pipeline bubbles,
\item to help out-of-order processors with speculative execution of branches: by
  hoisting expressions out of branches, code hoisting can effectively reduce
  the amount of code to be speculatively executed,
\item to decrease the number of live variables and thus to reduce register
  pressure,
\item to improve function inlining heuristics by reducing the size of a function body.
\end{itemize}

\begin{enumerate}
\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
\item O(n) Algorithm.
\item Hoisting equivalent computations in non-sibling BBs.
\item Hoisting across switch blocks.
\item Hoisting to improve PRE.
\item Hoisting to reduce critical path in OoO execution.
\item Ranking/Wavefront algorithm to hoist across switch blocks.
\end{enumerate}

\subsection{Enabling LICM by code-hoisting}

\begin{verbatim}
loop:
if (a)
  i1;
else
  i1;
\end{verbatim}

Here i1 is redundant and can be hoisted out of loop. But the LICM will
not do because it does not reason about instructions in the
conditional.  By code-hoisting i1 will be hoisted out of conditional
which will benefit the LICM.

\subsection{Improving the Inliner heuristics}
Reducing the number of instructions also reduces the inlining cost and
hence improves the inliner heuristics.

\subsection{Improving vectorization}
It will also benefit vectorizer by reducing the number
of basic blocks, should all the instructions in conditional get hoisted.

@@

@@

Code hoisting does not remove redundancies by itself but it enables
removal of redundancies by other passes.  e.g., Enable inlining.

@@

\subsection{Finding the candidates}
Finding if a set of instructions perform identical computations. Since this code
hoisting is based on GVN, we need the facility to compute GVN of instructions.
In llvm \cite{llvm}, GVN infrastructure is already set up and works well for
most kind of instructions except loads and store so we compute the GVN of loads
and stores separately. In order to value number stores, we value number (hash)
the address from where the value is to be loaded. For stores, we value number
(hash) the address as well as the value to be stored at that address.

@@

\subsection{Checking for legality}
Finding a hoisting point i.e., a common dominator of the set of instructions. If
such a dominator is found then checking if all the use-operands of the set of
instructions are available or not. In some cases if the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
passing the legality check. Checking that the side-effects of the computations
does not intersect with any side-effects between the instructions to be hoisted
and their hoisting point. For memory operations like loads, stores, calls etc.,
it is also required to check that all the paths from the hoisting point to the
end of the function should execute the exact instruction, in order to guarantee
correctness. Moreover, hoisting of memory operations is tricky on paths which
have indirect branch targets e.g., landing pad, case statements, goto labels
etc., because it becomes difficult to prove that all the paths from hoisting
point to the end of the function would execute the instruction. In our current
implementation we discard hoisting on such paths.

In the optimistic approach (Section-\ref{subsec:optimistic}), it is possible
that a common hoisting point of all the instructions is either too far away, or
not possible. In these cases, it is still possible to `partially' hoist a subset
of instructions by splitting the set of candidates and finding a closer hoisting
point for each subset. For more details see Section-\ref{subsec:partition}.

@@

\subsubsection{Hoisting loads}
The availability of operands to the load (an address) is checked at the hoisting
point. If it is not available we try to rematerialize the addresss if that is
possible.  Along the path, from current position of the load instruction through
the hoisting point, we find if there are aliasing stores. In those cases we
discard the candidate.

@@

\subsubsection{Hoisting calls}
For hoisting the call instructions, we check that the loads/stores in the call
does not create hazard with any loads/stores (in the caller) in the path from
callsite through the hoisting point, just like we did for loads and stores.

@@

\subsection{Making the transformation}
Once all the legality checks and profitability checks are satisfied, a copy of
the computation is inserted at the hoisting point and all the computations made
redundant by the new copy are removed. After that SSA semantics is restored by
updating the intermediate representation (IR) to reflect changes.

>>>>

This paper describes code hoisting, a technique that uses the information
computed for PRE to detect identical computations and has a transformation phase
whose goal differs from PRE: it removes identical computations from different
branches of execution.  These identical computations would not be redundant at
run-time and thus the number of run-time computations is not reduced.  Instead,
the goals of code hoisting are:
\begin{itemize}
\item to reduce the code size of the program;
\item to improve function inlining heuristics: functions become cheaper to
  inline by reducing their code size;
\item to expose more instruction level parallelism: by hoisting identical
  computations to be executed earlier, schedulers can move heavy computations
  earlier in order to avoid pipeline bubbles;
\item to help out-of-order processors with speculative execution of branches: by
  hoisting expressions out of branches, code hoisting can effectively reduce the
  amount of code to be speculatively executed and can reduce the critical path;
\item to reduce register pressure: by moving computations closer to the
  definitions of their operands;
\item to improve passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted and sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}

The complexity of code hoisting is linear in number of instructions that could
be hoisted in the program, matching the complexity of PRE on SSA form.  The
analysis phase is based on the Global Value Numbering (GVN), the same analysis
used for PRE, followed by the computation of a partition of identical
expressions to be hoisted in a same location to guarantee safety properties and
program performance, and followed by a simple code generation that adds the
identified instruction in the hoisting point and removes all the now redundant
expressions.

@@

\begin{itemize}
\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
\item Hoisting equivalent computations in non-sibling BBs.
\item Hoisting across switch blocks.
\item Ranking/Wavefront algorithm to hoist across switch blocks.
\end{itemize}

@@

@@

\subsection{Finding the candidates}
Finding if a set of instructions perform identical computations. Since this code
hoisting is based on GVN, we need the facility to compute GVN of instructions.
In llvm \cite{llvm}, GVN infrastructure is already set up and works well for
most kind of instructions except loads and stores so we compute the GVN of loads
and stores separately. In order to value number loads, we value number (hash)
the address from where the value is to be loaded. For stores, we value number
(hash) the address as well as the value to be stored at that address.

@@

\subsection{Checking for legality}
Finding a hoisting point i.e., a common dominator of the set of instructions. If
such a dominator is found then checking if all the use-operands of the set of
instructions are available or not. In some cases when the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
passing the legality check. Checking that the side-effects of the computations
does not intersect with any side-effects between the instructions to be hoisted
and their hoisting point. For memory operations like loads, stores, calls etc.,
it is also required to check that all the paths from the hoisting point to the
end of the function should execute the exact instruction, in order to guarantee
correctness. Moreover, hoisting of memory operations is tricky on paths which
have indirect branch targets e.g., landing pad, case statements, goto labels
etc., because it becomes difficult to prove that all the paths from hoisting
point to the end of the function would execute the instruction. In our current
implementation we discard hoisting through such paths.

In the optimistic approach (Section-\ref{subsec:optimistic}), it is possible
that a common hoisting point of all the instructions is either too far away, or
not legally possible. In these cases, it is still possible to `partially' hoist
a subset of instructions by splitting the set of candidates and finding a closer
hoisting point for each subset. For more details see
Section-\ref{subsec:partition}.

@@

\subsubsection{Hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction through
the hoisting point, we find if there are aliasing writes to memory. In those
cases we discard the candidate.

@@

\subsubsection{Hoisting calls}
For hoisting the call instructions, we check that the loads/stores in the call
would not hazard with any loads/stores (in the caller) along the path from the
callsite through the hoisting point, just like we did for loads and stores.

@@

\subsection{Making the transformation}
Once all the legality checks and profitability checks are satisfied for a set of
identical instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed. After that SSA semantics is restored by updating the
intermediate representation (IR) to reflect changes.
